Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-09-16 output/paper/paper.tex

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Umeå University,\href{mailto:joacim.rocklov@umu.se}{joacim.rocklov@umu.se }}\thanks{Heidelberg Institute of Global Health, University of Heidelberg}  \and
	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
}

\begin{document}
    \maketitle
    \begin{abstract}
        The new coronavirus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The SARS-CoV-2 pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}. Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers \citep{Jajosky2004_evaluation_reporting}. The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

Still, these unfinished statistics offer crucial information. If the pandemic is indeed slowing, we should not wait for the data to be finalized before using it. Rather, we argue that actual case counts and deaths should be nowcasted to account for reporting delay, thus allowing policymakers to use the latest numbers availiable without beinig misled by reporting bias.

Such predictions provide an additional feature that is perhaps even more important. They explicitly model the uncertainty about these unknown quantities, ensuring that all users of these data have the same view of the current state of the epidemic.

In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays. The prediction model is building on methods developed in ecology, referred to as the ``removal method'' \citep{Pollock1991_review_papers}.

To help motivate why such forecasting is needed, we now turn to the case of Sweden. The model is flexible by design, however, and could easily be applied to other countries as well.

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During a press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.

One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic \citep{Anderson2020_how_will}. In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response. However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces the illusion of a downward trend.

The death counts suffer from the longest reporting delay. In their daily press conference, the Swedish Public Health Agency warns for this by stopping the reported 7-day moving average trend line 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a sense of a downward trend. In fact, this might be the reason why the number of daily deaths have been underestimated repeatedly. At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 \citep{Ohman2020_antalet_virusdoda}.

\section{The removal method}
We propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}. However, the first refined mathematical treatment of the method is credited to \citet{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}. It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

The removal method that has three major advantages over simply reporting moving averages:

@@

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals \citep{Pollock1991_review_papers}. Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths
on a given day. Instead of traps we have the new reports of COVID--19 events. As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day. If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.

@@

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency\'s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. Note how the model predicts additional deaths above the moving average line.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/latest_prediction}
    \caption{Swedish Covid-19 deaths as of 2020-06-04 and model predictions}
    \label{fig:latest_prediction}
\end{figure}

\subsection{Model Performance}
To judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it is biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/lag_prediction_by_date}
    \caption{Model accuracy over time for four dates, compared to benchmark}
    \label{fig:four_dates}
\end{figure}

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \citep{bolin2019scale}. The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI\'s that cover the true value.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/model_metrics}
    \caption{Model accuracy over time for four dates, compared to a benchmark}
    \label{fig:model_metrics}
\end{figure}

@@

Whenever case statistics are collected from multiple sources and attributed to its actual event date in the middle of a public health emergency, similar reporting delays to the ones in Sweden will necessarily occur. The method described thus has implications and value beyond Sweden, for any situation where nowcasts of disease event frequencies are of relevance to public health.

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner. Given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a differewnt reporting structure than Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so.

@@

\section{Conclusion}
In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay. By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
\bibliographystyle{chicago}
\bibliography{bibliography}
\appendix
\section{Appendix}
\input{appendix.tex}%

>>>>

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
.
}

\begin{document}
    \maketitle
    \begin{abstract}
        The new coronavirus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely surveillance statistics, such as the trends in disease frequencies, are a key to evaluating and planning responses. But due to reporting delays, the most recently reported numbers are often underestimating of the total number of new infections, hospitalizations and deaths. This sometimes creates an illusion of a downward trend. Here we describe a statistical methodology for predicting the expected actual daily surveillance frequencies and their uncertainty, considering historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The SARS-CoV-2 pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, surveillance data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}. Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers to evaluate and plan rapid responses \citep{Jajosky2004_evaluation_reporting}. But as a general rule, the latest data are often not complete, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

Still, these unfinished surveillance statistics offer crucial information. If the pandemic is indeed slowing, we should not wait for the data to be finalized before using it. Rather, we argue that actual case counts and deaths should be nowcasted to account for reporting delays, thus allowing policymakers to use the latest numbers available surveillance without being misled by the obvious reporting bias.

Such predictions provide an additional feature that is perhaps even more important. They explicitly model the uncertainty in these unknown quantities, ensuring that all users of these data have the same view of the current state of the epidemic, including the uncertainty.

In this paper we describe a statistical methodology for nowcasting the surveillance statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays. The prediction model is building on methods developed in ecology, referred to as the ``removal method'' \citep{Pollock1991_review_papers}.

To help motivate why such forecasting is needed, we now turn to the case of Sweden, and illustrate how this method and can provide more reliable and relevant data to decision-makers in the midst of a pandemic. We stress, however, that the model is flexible by design and that it could easily be applied to other countries and surveillance data as well.

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During frequently re-occurring press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.

One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic \citep{Anderson2020_how_will}. In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response. However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces makes the recent data presented less useful.

The death counts strongly suffer from the longer reporting delays. In their daily press conference, the Swedish Public Health Agency warns for this by stopping the reported 7-day moving average trend line 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a false impression of sense of a downward trend. In fact, in the Swedish situation, this might be the reason why the number of daily deaths have been underestimated repeatedly. As an example illustrating this, at the peak of the Swedish outbreak in the spring 2020, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 \citep{Ohman2020_antalet_virusdoda}.

\section{The removal method}
To correct for the reporting delays we propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}. However, the first refined mathematical treatment of the method is credited to \citet{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}. It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

The removal method has three major advantages over the simplistic approach to just reporting moving averages:

@@

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals \citep{Pollock1991_review_papers}. Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths on a given day. For COVID-19, instead of traps we have the new reports of COVID--19 events. As the number of non-yet reported deaths for a given day declines according to some patterns over time, we can use this patterns to draw inference on how many actually died a specific day. If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number, not the reported one.

@@

We note that a similar method has been proposed for delay adjusted reporting, but without using the removal method \cite{lawless1994adjustments}. In \cite{lawless1994adjustments} the author developed the Occurred But Not Reported method (OBNR), which can give a frequentist point estimate of the number of deaths, and under a Poisson distributional assumption, generate prediction intervals. A Bayesian version of this method for nowcasting was developed in \cite{hohle2014bayesian}.

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency\'s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. We note specifically how the model predicts additional deaths above the moving average line.

\subsection{Model Performance}
To judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it is biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by using a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated.
In forecast evaluation, a function that guarantees, asymptotically, that the forecasting distribution (if available) will be chosen, is known as a proper scoring rule, see \cite{gneiting2007strictly}.
In \cite{hohle2014bayesian} validation was done using the proper scoring rules logarithmic score and ranked probability score. These two scores requires access to either the PMF or CDF of the predictive distribution. Since we don't have easy access to these functions we instead use the  SCRPS (Scaled continuous proability ranking score, see definition in Appendix) \citep{bolin2019scale}. The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI\'s that cover the true value.

@@

@@

Whenever case statistics are collected from multiple sources and attributed to its actual event date in the middle of a public health emergency, similar reporting delays to the ones in Sweden will necessarily occur. The method described thus has implications and value beyond Sweden, for any situation where nowcasts of surveillance data are of relevance to public health.

In \cite{schneble2020nowcasting} nowcasting of number of Covid-19 deaths in Germany was preformed. For nowcasting they applied the OBNR method, using penalized smoothing splines. The accuracy of their nowcasting model is, however, not possible to compare to our model as the study by Schneble et al. didn\'t validate the nowcasting on data withheld from the model fitting.

Our method has its limitations. As presented, the model assumes that all deaths are reported in the same manner. Given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so.

@@

\section{Conclusion}
In this paper, we provide a method to accurately adjust for reporting delays. This is illustrated by nowcasting daily Covid-19 deaths in Sweden, which are reported with strong delay patterns. By systematically accounting for the delay by nowcasting, better timing in the evaluation and planning of responses and policy can be achieved, which is of great importance for in pandemic situations. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during pandemics.

\section{Additional Information}
The authors declare no conflict of interest.
%%%%%%%%%%%%%%%%%%%%%%%%%%%'

%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

% Author contributions
\bibliographystyle{chicago}
\bibliography{bibliography}
\input{contributions.tex}%

\newpage

\input{Figure1.tex}%
\newpage
\input{Figure2.tex}%
\newpage

\input{Figure3.tex}%

\newpage
\appendix
\section{Appendix}
\input{appendix.tex}%
